The Design and Implementation of the Synchronous Language \CEU

### 2016-03-28 paper.tex
...

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
Both problems have an extensive coverage and debate in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of the loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to share memory and interact with stateful system calls safely.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward single-threaded implementation targeting resource-constrained 
systems.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C via function annotations and enforced 
finalization for external resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU is a synchronous reactive language inspired by Esterel with support for 
multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events, which are broadcast to all awaiting trails.
By synchronous, we mean that trails at any given moment are either reacting to 
the current event or are awaiting another event;
in other words, trails never react to different events simultaneously.

In the sections that follow, we discuss the main differences between \CEU and 
Esterel:
unique and queue-based external events (Section~\ref{sec.ceu.exts}),
stack-based internal events (Section~\ref{sec.ceu.ints}),
static temporal analysis for concurrent statements (Section~\ref{sec.ceu.analysis}),
safe integration with \C (Section~\ref{sec.ceu.c}), and
first-class synchronized timers (Section~\ref{sec.ceu.timers}).

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel (a) and \CEU (b) for the following control 
specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

\CEU employs the synchronous model, in which programs advance in a sequence of 
discrete reactions to external events.
%
It also has a strong imperative flavor, with explicit control flow through 
sequences, loops, and parallels, and also assignments.
%
Being designed for control-intensive applications, \CEU provides support for 
concurrent lines of execution and broadcast communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

@@

@@

Esterel defines a logical unit of time as a discrete sequence of instants or 
``ticks''.
At each tick, depending on external stimuli the environment provides, an 
arbitrary number of input events can be active simultaneously (including zero).
The set of input events awakes the program, which must react within the current 
tick, in bounded time.

In \CEU, each unique external stimulus itself defines a unit of time.
The execution model for \CEU programs is as follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react completely to an occurring event completely before 
handling the next.
%
Based on the synchronous hypothesis, a program takes no time on step 2 and is 
always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running (step 2), it is enqueued to run in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
\CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
Since Esterel and \CEU define discrete logical units of time, we describe how 
the events (which are the same in all boxes) fit into their timelines.

\code{Box-1} uses fixed-length ticks in Esterel, within which reactions to 
occurring inputs have to fit~\cite{esterel.fixed} .
%
Here, we assume a \code{R(boot)} reaction which happens before any input at 
\code{tick-0}.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is logical and discrete, the event is delayed to the next tick.
%
Note that the reaction \code{R(A)} takes more time than \code{tick-1}, causing 
a \emph{timing violation} in the program execution (after the end of 
\code{tick-1} and before the end of \code{R(A)}).
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occurred during \code{tick-2}, the CPU stays idle during 
the whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur ``simultaneously'' during the idle \code{tick-3}.
However, because the program cannot detect both occurrences of \code{E}, only 
one is considered in \code{R(D+E)}.

\code{Box-2} is an alternative that allows variable-length ticks in Esterel 
that adjust to the time the corresponding reaction takes to 
complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also yields
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
start reaction \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}.

\code{Box-3} illustrates the unique and queue-based policy for input events in 
\CEU.
We also assume a \code{R(boot)} reaction before any input.
%
Because the event \code{A} is unique during \code{tick-0}, the behavior in \CEU 
is similar to \code{Box-2} for the first two units of time.
%
However, the events \code{B} and \code{C} are not simultaneous in \CEU and are 
handled in subsequent reactions \code{R(B)} and \code{R(C)}.
%
We assume that the CPU time for \code{R(B+C)} in Esterel is roughly the same as 
\code{R(B)+R(C)} in \CEU.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.

We decided for the unique and queue-based semantics for the reasons that 
follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but explicitly defined for each use case.
Note that in the case of Esterel, simultaneity depends on the imprecise 
definition of ticks.
For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
simultaneous, even though \code{A} and \code{B} actually happen much closer to 
one another.
After introducing internal events and first-class timers, we show an example 
that detects simultaneous button clicks in Section~\ref{sec.ceu.discussion}.
%
\item \textbf{Unique input events imply mutual exclusion:}
Because they are atomic, reactions to different events never overlap.
The possibility to reason about each input individually is a prerequisite for 
the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
\end{itemize}

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the system will accumulate delays between the real occurrence and 
actual reaction to events.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
will accumulate for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO: Such analsysis is orthogonal

@@

>>>>

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis such as to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
Both problems have an extensive coverage and debate in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of the loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to interact with stateful system calls safely and makes shared-memory 
concurrency not as straightforward as reading and writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis to 
improve safety:
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel (a) and 
\CEU (b) for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel:
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events (including zero), depending on external stimuli the environment 
provides.
In contrast, \CEU defines time as a discrete sequence of reactions to
external stimuli.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The execution of a program in \CEU is as follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes no time on step 2 and is 
always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running (step 2), it is enqueued to run in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
\CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
Esterel and \CEU use discrete logical units of time in which the occurring 
events (which are the same in all boxes) have to fit somehow:
%
\begin{itemize}
\item \code{Box-1}:
Esterel with fixed-length ticks,
within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
%
We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes inthe next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur ``simultaneously'' during the idle \code{tick-3}.
However, the program can only detect one occurrence of \code{E}, which is 
considered in \code{R(D+E)}.
%
\item \code{Box-2}:
Esterel with variable-length ticks that adjust to the CPU time the 
corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
start reaction \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{Box-3}:
\CEU with unique and queue-based policy for input events.
We also assume a \code{R(boot)} reaction before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two units of time 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} simultaneous, 
which are handled in subsequent reactions \code{R(B)} and \code{R(C)}.
%
We assume that the CPU time for \code{R(B+C)} in Esterel is roughly the same as 
\code{R(B)+R(C)} in \CEU.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
events \code{B} and \code{C} are simultaneous, even though \code{A} and 
\code{B} actually happen much closer to one another.
In Section~\ref{sec.ceu.discussion}, after introducing internal events and 
first-class timers, we show an example that detects simultaneous button clicks.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about the program and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the application continuously accumulates delays between the real 
occurrence and actual reaction to events.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
accumulates for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.
% TODO: Such analsysis is orthogonal

@@

TODO: queue-based/unique in the semantics of the language and not in the 
library

permeates all aspects
